Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-21 co-regular-EPQ.tex
introduction
modify
small things

Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions which contains only one maximal priority. This is possible because any data-differentiated execution for which $\Gamma(e)$ is false has such a projection.
%Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions where any two occurring priorities are comparable (w.r.t. $\prec$). This is possible because any data-differentiated execution which is not $\Gamma$-linearizable has such a projection.
Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities comparable to $p$. Then,

@@

\begin{figure}[htbp]
TODO UPDATE THE ARGUMENTS OF ACTIONS
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that is not $\mathsf{MatchedMaxPriority}^{>}$-linearizable}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

@@

The execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} is not $\mathsf{MatchedMaxPriority}^>$-linearizable because the left-right constraint of the maximal priority value $b$ contains a cycle: $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$. We prove that the presence of such a cycle is equivalent to the execution not being $\mathsf{MatchedMaxPriority}^>$-linearizable (see Appendix~\ref{}).
%The existence of cycle though value with maximal priority in left-right constraint formally modelled the phenomenon in \figurename~\ref{fig:introduce gap for EPQ1Lar}. For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$ is a cycle in left-right constraint. We then prove that getting rid of cycle though value with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$.
When the left-right constraint of the maximal priority value $x$ contains a cycle of the form $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ for some $d_1$,$\ldots$,$d_n\in \mathbb{D}$, we say that $x$ is \emph{covered} by $d_1,\ldots,d_m$. The shape of such a cycle can be detected using our class of automata, the only complication being the unbounded number of values $d_1$,$\ldots$,$d_n$. However, by data independence, whenever an implementation contains such an execution it also contains an execution where all the values $d_1$,$\ldots$,$d_n$ are renamed to the same value $a$, and $x$ is renamed to $b$. Therefore, our automata can be defined over a fixed set of values $a$, $b$, and $\top$ (for the operations outside of the non-linearizable projection).

@@

{\color {blue} Assume that $x$ is covered by $d_1,\ldots,d_m$, we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other value into $c$ by data-independence. Since there are four cases of projection of execution to $x$ when their is a cycle through $x$. To ease the reading, for each of such case we generate a register automata, and then let the $\mathsf{MatchedMaxPriority}^{>}$-complete automaton be the union of the four register automata. For example, for the case when the projection of $e$ to $x$ is as in \figurename~\ref{fig:introduce gap for EPQ1Lar} (rename $b$ into $x$), we construct a register automaton $\mathcal{A}_{\textit{l-lar}}^1$ in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. Here transition labels $c_1 = c + \textit{ret}(\textit{rm},a)$, $c_2 = c + \textit{call}(\textit{put},a,=r)$, $c_3 = c_2 + \textit{ret}(\textit{rm},a)$, where $c = \textit{call}(\textit{put},d,\top),\textit{ret}(\textit{put},d,\top), \textit{call}(\textit{rm},d)$, $\textit{ret}(\textit{rm},d),\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty})$. There are three paths in \figurename~\ref{fig:introduce gap for EPQ1Lar} from $q_{\textit{init}}$ to $q_7$, and they comes from that there are three possible positions for the first $\textit{call}(\textit{put},a,\_)$ in the captured execution, which is shown in \figurename~\ref{fig:executions APQ1Lar-1 in paper}. In \figurename~\ref{fig:automata APQ1Lar-1 in paper}, the paths $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_7$, $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_{10}$ and $q_1 \rightarrow q_9 \rightarrow q_{10} \ldots \rightarrow q_7$ corresponds to the first, second and third cases in \figurename~\ref{fig:executions APQ1Lar-1 in paper}, respectively.
}

>>>>

Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions which contain only one maximal priority. This is possible because any data-differentiated execution for which $\Gamma(e)$ is false has such a projection.
%Also, we consider $\Gamma$-complete automata for $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$, recognizing executions where any two occurring priorities are comparable (w.r.t. $\prec$). This is possible because any data-differentiated execution which is not $\Gamma$-linearizable has such a projection.
Formally, given a data-differentiated execution $e$ and $p$ a maximal priority in $e$, $e\vert_{\preceq p}$ is the projection of $e$ to the set of values with priorities comparable to $p$. Then,

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that is not $\mathsf{MatchedMaxPriority}^{>}$-linearizable. We represent each operation as a time interval whose left, resp., right, bound corresponds to the call, resp., return action. Operations adding and removing the same value are aligned vertically.}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

@@

The execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} is not $\mathsf{MatchedMaxPriority}^>$-linearizable because the left-right constraint of the maximal priority value $b$ contains a cycle: $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$. We prove that the presence of such a cycle is equivalent to the execution not being $\mathsf{MatchedMaxPriority}^>$-linearizable (see Appendix~\ref{}).
%The existence of cycle though value with maximal priority in left-right constraint formally modelled the phenomenon in \figurename~\ref{fig:introduce gap for EPQ1Lar}. For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$ is a cycle in left-right constraint. We then prove that getting rid of cycle though value with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$.
When the left-right constraint of the maximal priority value $x$ contains a cycle of the form $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ for some $d_1$,$\ldots$,$d_n\in \mathbb{D}$, we say that $x$ is \emph{covered} by $d_1,\ldots,d_m$. The shape of such a cycle (i.e., the alternation between call/return actions of $\textit{put}$/$\textit{rm}$ operations) can be detected using our class of automata, the only complication being the unbounded number of values $d_1$,$\ldots$,$d_n$. However, by data independence, whenever an implementation contains such an execution it also contains an execution where all the values $d_1$,$\ldots$,$d_n$ are renamed to the same value $a$, and $x$ is renamed to $b$. Therefore, our automata can be defined over a fixed set of values $a$, $b$, and $\top$ (for the operations outside of the non-linearizable projection).

@@

{\color {blue} Assume that $x$ is covered by $d_1,\ldots,d_m$, we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other value into $c$ by data-independence. Since there are four cases of projection of execution to $x$ when their is a cycle through $x$. To ease the reading, for each of such case we generate a register automata, and then let the $\mathsf{MatchedMaxPriority}^{>}$-complete automaton be the union of the four register automata. For example, for the case when the projection of $e$ to $x$ is as in \figurename~\ref{fig:introduce gap for EPQ1Lar} (rename $b$ into $x$), we construct a register automaton $\mathcal{A}_{\textit{l-lar}}^1$ in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. Here transition labels $c_1 = c + \textit{ret}(\textit{rm},a)$, $c_2 = c + \textit{call}(\textit{put},a,=r)$, $c_3 = c_2 + \textit{ret}(\textit{rm},a)$, where $c = \textit{call}(\textit{put},\top,\textit{true}),\textit{ret}(\textit{put},\top,\textit{true}), \textit{call}(\textit{rm},d)$, $\textit{ret}(\textit{rm},d),\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty})$. There are three paths in \figurename~\ref{fig:introduce gap for EPQ1Lar} from $q_{\textit{init}}$ to $q_7$, and they comes from that there are three possible positions for the first $\textit{call}(\textit{put},a,\_)$ in the captured execution, which is shown in \figurename~\ref{fig:executions APQ1Lar-1 in paper}. In \figurename~\ref{fig:automata APQ1Lar-1 in paper}, the paths $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_7$, $q_1 \rightarrow q_2 \rightarrow q_3 \ldots \rightarrow q_{10}$ and $q_1 \rightarrow q_9 \rightarrow q_{10} \ldots \rightarrow q_7$ corresponds to the first, second and third cases in \figurename~\ref{fig:executions APQ1Lar-1 in paper}, respectively.
}
